Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-21 paper.qmd
Add intro to application data sets

library(Rtsne)
library(umap)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)
library(grid)
set.seed(20240110)

@@

In this article, mainly focus on five NLDR techniques. They are tSNE, UMAP, PHATE, TriMAP [@article02], and Pairwise Controlled Manifold Approximation (PaCMAP) [@article01].

@@

PaCMAP [@article01] is different because it adds supervised learning to make a 2D representation while keeping the relationships between pairs of points. It builds a graph using distances between pairs and then makes the 2D representation better using a customizable loss function. What's special about PaCMAP is that it can use class labels or extra information to guide how it makes the 2D representation. This gives users a way to change how PaCMAP works to fit their needs better.

@@

Typically, the MSE plot tends to have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or levesl off, indicating that a higher number of bins generates a smaller error. This MSE plot is useful in determining the effective number of bins required to construct the 2D model. @fig-diagnosticpltScurve (b) has created a graph that shows how the Mean Squared Error (MSE) changes depending on the number of bins assigned in the hexagonal grid for UMAP applied to the S-curve dataset. As per the rule of thumb mentioned earlier, the effective number of bins is determined to be $84$, with $6$ bins along the x-axis and $14$ bins along the y-axis. Furthermore, @fig-modelScurve shows the 2D model constructed using UMAP with the selected total number of bins and how the model overlay on high-D data.

@@

## To draw with AIC
aic_plot <- ggplot(mse_df, aes(x = num_bins, y = aic
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 84, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

@@

aic_plot + mse_plot +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

```{r msebinsnldr}
#| echo: false
#| message: false
#| warning: false

# num_bins_for_nldr_methods_df <- tibble::tibble(NLDR_technique = "tSNE", tot_bins = 72, aic = -21882.462)
# 
# num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
#   tibble::add_row(NLDR_technique = "UMAP", tot_bins = 98, aic = -20013.266)
# 
# num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
#   tibble::add_row(NLDR_technique = "PHATE", tot_bins = 90, aic = -15036.91)
# 
# num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
#   tibble::add_row(NLDR_technique = "TriMAP", tot_bins = 154, aic = -23001.90)
# 
# num_bins_for_nldr_methods_df <- num_bins_for_nldr_methods_df |> 
#   tibble::add_row(NLDR_technique = "PaCMAP", tot_bins = 132, aic = -21330.059)
# 
# # knitr::kable(
# #   num_bins_for_nldr_methods_df,
# #   col.names = c('NLDR technique', 'total number of bins'), format = "latex", 
# #              caption = "Total number of bins suggested by rule of thumb for each NLDR technique applied to S-curve dataset."
# # ) |> 
# #   kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
# #   row_spec(0, bold = TRUE)
# num_bins_for_nldr_methods_df$tot_bins <- as.character(num_bins_for_nldr_methods_df$tot_bins)
# colnames(num_bins_for_nldr_methods_df) <- c('NLDR technique', 'total number of bins', "AIC")
# num_bins_for_nldr_methods_df <- tibble::remove_rownames(num_bins_for_nldr_methods_df)
# 
# xtable(num_bins_for_nldr_methods_df, caption = "Total number of bins suggested by rule of thumb and AIC values for each NLDR technique applied to S-curve dataset.", label = "tbl-msebinsnldr") |>
#   xtable2kable() |> 
#   kable_styling(latex_options = c("striped", "HOLD_position"), position = "center", full_width = FALSE) |>
#   row_spec(0, bold = TRUE) |>
#   row_spec(4, bold=T, color = "red")

<!--Adjusting the parameter $b_1$ provides control over the total number of bins $b$, allowing us to fine-tune the hexagonal grid.-->

<!--first value-->

<!--                  
```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false
hex_full_count_df_loop1 <- read_rds("data/s_curve/s_curve_hex_3.rds")

p1 <-  ggplot(data = hex_full_count_df_loop1, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "a", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false
hex_full_count_df_loop2 <- read_rds("data/s_curve/s_curve_hex_11.rds")

p2 <-  ggplot(data = hex_full_count_df_loop2, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "b", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false
hex_full_count_df_loop3 <- read_rds("data/s_curve/s_curve_hex_20.rds")

p3 <-  ggplot(data = hex_full_count_df_loop3, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = polygon_id, fill = std_counts)) +
  geom_point(data = UMAP_data, aes(x = UMAP1, y = UMAP2), alpha = 0.5) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  coord_cartesian(xlim =c(-5, 7), ylim = c(-10, 10)) +
  theme_void() +
  theme(legend.position="none", legend.direction="horizontal", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=8), #change legend title font size
        legend.text = element_text(size=6)) +
  guides(fill = guide_colourbar(title = "Standardized count")) +
  annotate(geom = 'text', label = "c", x = -Inf, y = Inf, hjust = -0.3, vjust = 1, size = 3)

```{r}
#| eval: false
#| echo: false
#| label: fig-binsize1
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the **UMAP** projections of **S-curve** training data: (a) b = 32 (4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b = 840 (21, 40), s = 1.643542. The hexbins are colored based on the density of points, with darker colors indicating higher point density and yellow color representing lower point density within each bin. What is the number of bins that would be effective in representing low-dimensional data?"

```{r}
#| warning: false
#| echo: false
#| eval: false

data <- read_rds("data/s_curve/s_curve.rds")

shape_value_curve <- calculate_effective_shape_value(.data = UMAP_data,
                                                   x = UMAP1, y = UMAP2)

num_bins_vec <- 1:13 ## Number of bins along the x-axis

vec <- stats::setNames(rep("", 5), c("number_of_bins", "number_of_observations", "total_error", "total_mse", "num_bins_x"))  ## Define column names

  model_object <- fit_high_d_model(training_data = training_data, nldr_df_with_id = UMAP_data, x = "UMAP1", y = "UMAP2", num_bins_x = num_bins_vec[i], shape_val = shape_value_curve,
                             is_bin_centroid = TRUE,
                             is_rm_lwd_hex = FALSE,
                             benchmark_to_rm_lwd_hex = NA,
                             is_avg_high_d = TRUE, column_start_text = "x")

  pred_df_training <- predict_2d_embeddings(test_data = training_data, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, type_NLDR = "UMAP")

  # pred_df_training <- predict_hex_id(df_bin_centroids = centroid_df_training, nldr_df_test = UMAP_data, x = "UMAP1", y = "UMAP2")

  eval_df_training <- generate_eval_df(data = data, prediction_df = pred_df_training, df_bin_centroids = centroid_df_training, df_bin = avg_df_training, col_start = "x")

MSE_df_umap_s_curve <- eval_data_training |>
  mutate(data_type = "training",
         method = "UMAP")
```

```{r}
#| warning: false
#| echo: false
#| eval: false

## To draw with AIC
aic_plot <- ggplot(MSE_df_umap_s_curve |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 144, linetype="dashed", 
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot <- ggplot(MSE_df_umap_s_curve |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                       y = total_mse,
                                       color = method
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 144, linetype="dashed", 
                color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

@@

After setting up the hexagonal grid with an appropriate number of bins, it is possible that some hexagonal bins may have very few or no data points within them. To ensure comprehensive coverage of the NLDR data, it is necessary to select hexagonal bins that have a considerable number of data points within their hexagons. To achieve this, the standard number of points within each hexagon is first calculated. This standard count is obtained by dividing the number of points within each hexagon by the maximum number of points in the grid (see @eq-equationp2). Next, the bins that have a standard number of points less than a certain benchmark value are removed. After removing the hexagons that have insufficient data density, the hexagons with more substantial data representation are used to construct the 2D model.

@@

Furthermore, it is crucial to choose the benchmark value to remove low-density hexagons carefully. If we remove unnecessary bins, then it may result in long edges and an uneven 2D model. Therefore, instead of removing hexagons only identified by the benchmark value, we examine the standard number of points in the neighboring hexagons of the identified low-density hexagons. If the neighboring bins also have low counts, then only those bins will be removed. The other identified bins will be kept and used for constructing the 2D model along with the high-density bins (see Appendix for more details).

The benchmark value for removing low-density hexagons has a range of $0$ and $1$. When analyzing how these benchmark values affect model performance, it's important to observe the change in Mean Squared Error (MSE) as the benchmark value increases. The MSE exhibits a gradual increase as the benchmark value progresses from $0$ to $1$. Assessing this rate of increase is crucial. If the increment is not substantial, the decision might lean towards retaining low-density hexagons. According to @fig-diagnosticpltScurvelwd, the change in MSE for NLDR techniques is relatively small across different benchmark values, except for the PHATE method.

Consequently, it is not necessary to remove low-density hexagons when constructing the 2D model for the S-curve dataset in NLDR techniques, except for PHATE. While there may be some fluctuations in MSE between benchmark values of 0 to 1, the absence of significant changes suggests that maintaining low-density hexagons is preferable. However, to choose the best selection, it's recommended to examine the benchmark values around the first local minimum in the MSE curve. However, the MSE for PHATE fluctuates when benchmark values range between 0 and 1. Therefore, it is necessary to explore the benchmark values surrounding the first local minimum to determine an effective benchmark value. There is no universal rule for selecting a process. As shown in @fig-diagnosticpltScurvelwd, the MSE values for PHATE fluctuate, and the first local minimum is at $0$. Therefore, nearby benchmark values such as $0.01$, $0.04$, and $0.06$ were explored to identify the best value for preserving the data structure. In this case, a benchmark value of $0.06$ was chosen. The resulting 2D model is shown in @fig-modelScurve (c), while the video link provides a visual representation of how the model appears in high-D.

@@

@@

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
pcentroid_plot_s_curve <- ggplot(data = hex_grid, aes(x = x, y = y)) +
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = df_bin_centroids, aes(x = c_x, y = c_y), color = "#33a02c") +
  coord_fixed() +
  theme_linedraw() +
  theme(legend.position = "bottom", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "(a), (b) 2D model."
#| label: fig-modelScurve

@@

@@

@@

To assess how well our method captures and represents the underlying structure of the high-D data, Mean Squared Error (MSE) and Akaike Information Criterion (AIC) are used. When computing MSE, total model error (see @sec-summary) is divided by the number of observations to make it as a mean value (see @eq-equation9).

@@

Furthermore, to incorporate with the parameters in our model, which is the non-empty bins, and as a statistical measure used for model selection and comparison, AIC is used (see @eq-equation10).

$$
\text{AIC} = 2b'p + np * log(\text{MSE})
$$ {#eq-equation10}

Here, $b'$ signifies the number of bins non-empty bins, $p$ denotes the number of dimensions in the high-D data, and $n$ represents the number of observations.

@@

phate_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = PHATE_data_gau, x = "PAHTE1", y = "PAHTE2"))) |>
  dplyr::rename(c("PAHTE1" = "scaled_PAHTE1", 
                  "PAHTE2" = "scaled_PAHTE2")) |>
  dplyr::mutate(ID = 1:NROW(PAHTE_data_gau))

@@

@@

In the field of single-cell studies, a common analysis task involves clustering to identify groups of cells with similar expression profiles. Analysts often turn to NLDR techniques to verify and identify these clusters and explore developmental trajectories. In clustering workflows, the main objective is to verify the existence of clusters and subsequently identify them as specific cell types by examining the expression of "known" marker genes. Marker genes are specific genes or transcripts that are uniquely or highly expressed in particular cell types. Marker genes are identified through differential gene expression (DEG) analysis, comparing the expression levels of genes across different cell types. In this context, a "faithful" embedding should ideally preserve the topology of the data, ensuring that cells corresponding to the same cell type are situated close to the high-dimensional space.

### Processing steps

To begin our analysis, we installed the Peripheral Blood Mononuclear Cells (pbmc) data set obtained from 10x Genomics using the `SeuratData` R package [@Rahul2019], which facilitates the distribution of data sets in the form of Seurat objects [@Yuhan2021]. This data set contains 13,714 features (genes) across 2,700 samples (cells) within a single assay. The active assay is RNA, with 13,714 features representing different gene expressions. Then, the cells that have unique feature counts over 2,500 or less than 200 and cells that have > 5\% mitochondrial counts are filtered. After removing unwanted cells from the dataset, the next step is to normalize the data. By default, we employ a global-scaling normalization method “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result.

Next, top 1000 genes are selected by Festem [@Chen2023] DEG method. Then, scale the data. Then, perform PCA on the scaled data. Louvain algorithm is used to cluster the single cells based on the genes selected by Festem. Using the Festem-selected genes and 15 PCs, identified 10 clusters in the PBMC3k data and annotated them based on the expression of canonical markers. The 10 clusters included immune cells such as naive CD4 T cells, memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes, Natural Killer (NK) cells, B cells and Dendritic Cells (DC). In addition to these common cell types, Festem identified a fine cell type, CD27− CD4+ memory T cells, which were often missed by other methods.

- Human peripheral blood mononclear cells (PBMCs) data

- Intro to the data set

Used Louvain algorithm to cluster the single cells based on the genes selected by Festem method

- Explain what "Evaluate Festem on DEG detection" means, in terms of this data

- How did they "identified 10 clusters"? What were the variables used, PCs or full data or 2D representation, spell this out.

- Used top 1000 genes selected by Festem to run PCA, top 15 PCs (original paper mentioned)

- The "10 clusters included immune cells such as naive CD4 T ..." does this mean that every cluster had all of these, or one cluster was mostly "CD4 T cells", another mostly "memory CD4 T", ...?

- How is your use of this data different from the original application.

- Original paper used this data set to compare different DEG methods, but we need to look how the performance of UMAP with the author's selected parameter choice.

- Why 15 PCs? Is that from the original paper? YES

@Chen2023

- Intro to the data set (This data set contains 13,714 features across 2,700 samples within a single assay. The active assay is RNA, with 13,714 features representing different gene expressions.)

- 2622 cells

- Evaluate Festem on DEG detection (Festem enabled identification of often-missed fine cell types)

- Using the Festem-selected genes, identified 10 clusters in the PBMCsk data and annotated them based on the expression of canonical markers

- The 10 clusters included immune cells such as naive CD4 T cells, memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes, Natural Killer (NK) cells, B cells and Dendritic Cells (DC).

- In addition to these common cell types, Festem identified a fine cell type, CD27− CD4+ memory T cells, which were often missed by other methods.

- The CD27. CD4+ memory T cells identified by Festem expressed common marker genes (IL7R
and S100A4) of memory T cells, but did not express CD27. These cells also had downregulated
expression of SELL, CCR7, MAL and LEF1, and upregulated expression of CCL5 (Fig. 4B) and
thus were the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+ memory T
cells were known to be at a more differentiated state and have stronger antigen-recall responses
than their CD27+ counterparts.

- 15 PCs

- Num bins along the x-axis = 23, shape parameter = 0.8772751

- learned from the model: There are 3 well-separated clusters in 2D, but if look at the model in high-D space, the three clusters are much closer. Also, there is some continuity within the clustering structure and it didn't capture well. Therefore UMAP with n:neighbour 30 is not a good representation for this data set.

- The most important parameter is n_neighbors - the number of approximate nearest neighbors used to construct the initial high-dimensional graph. It effectively controls how UMAP balances local versus global structure - low values will push UMAP to focus more on local structure by constraining the number of neighboring points considered when analyzing the data in high dimensions, while high values will push UMAP towards representing the big-picture structure while losing fine detail.

- The second parameter we’ll investigate is min_dist, or the minimum distance between points in low-dimensional space. This parameter controls how tightly UMAP clumps points together, with low values leading to more tightly packed embeddings. Larger values of min_dist will make UMAP pack points together more loosely, focusing instead on the preservation of the broad topological structure.

- The third parameter is metric. This parameter plays a crucial role in determining how distances are computed within the ambient space of the input data (The term "ambient space" refers to the original or initial space in which the data exists before any transformation or dimensionality reduction).

- If it is completely a mess, then that is also further evidence that the fit is poor in high-dimensions.

<!--https://dicook.github.io/mulgar_book/7-spin-and-brush.html#fig-penguins-bs-detourr-->

>>>>

library(Rtsne)
library(uwot)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)
library(grid)
set.seed(20240110)

@@

In this article, mainly focus on five NLDR techniques. They are tSNE, UMAP, PHATE, TriMAP [@article02], and Pairwise Controlled Manifold Approximation (PaCMAP) [@Yingfan2021].

@@

PaCMAP [@Yingfan2021] is different because it adds supervised learning to make a 2D representation while keeping the relationships between pairs of points. It builds a graph using distances between pairs and then makes the 2D representation better using a customizable loss function. What's special about PaCMAP is that it can use class labels or extra information to guide how it makes the 2D representation. This gives users a way to change how PaCMAP works to fit their needs better.

@@

Typically, the MSE plot tends to have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or levesl off, indicating that a higher number of bins generates a smaller error. This MSE plot is useful in determining the effective number of bins required to construct the 2D model. @fig-diagnosticpltScurve has created a graph that shows how the Mean Squared Error (MSE) changes depending on the number of bins assigned in the hexagonal grid for UMAP applied to the S-curve dataset. As per the rule of thumb mentioned earlier, the effective number of bins is determined to be $84$, with $6$ bins along the x-axis and $14$ bins along the y-axis. Furthermore, @fig-modelScurve shows the 2D model constructed using UMAP with the selected total number of bins and how the model overlay on high-D data.

@@

## To draw with AIC
# aic_plot <- ggplot(mse_df, aes(x = num_bins, y = aic
# )) +
#   geom_point() +
#   geom_line() +
#   geom_vline(xintercept = 84, linetype="solid",
#                 color = "red", size=0.8, alpha = 0.5) +
#   theme_light() +
#   theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
#         axis.title = element_text(size = 7),
#         axis.text = element_text(size = 7)) +
#   #scale_colour_manual(values = c("#377eb8")) +
#   ylab("AIC") +
#   xlab("total number of bins")f
## Effective number of bins along x-axis

@@

mse_plot
```

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_s_curve$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
pcentroid_plot_s_curve <- ggplot(data = hex_grid, aes(x = x, y = y)) +
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = df_bin_centroids, aes(x = c_x, y = c_y), color = "#33a02c") +
  coord_equal() +
  theme_linedraw() +
  theme(legend.position = "bottom", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

cell_count_plot <- ggplot(df_bin_centroids_s_curve, aes(x = reorder(as.factor(hexID), -std_counts), y = std_counts)) +
  geom_quasirandom() + xlab("hexagonal id") + ylab("Standardized cell count") +
  geom_hline(yintercept = 0.255, colour = "#de2d26") +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 7),
        axis.text.x = element_text(angle = 90))

distance_plot <- plot_dist(distance_s_curve) +
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12))
# 
# distance_plot

@@

After setting up the hexagonal grid with an appropriate number of bins, it is possible that some hexagonal bins may have very few or no data points within them. To ensure comprehensive coverage of the NLDR data, it is necessary to select hexagonal bins that have a considerable number of data points within their hexagons. To achieve this, the number of points within each hexagon is first calculated. Then, the standard count is computed by dividing the number of points within each hexagon by the maximum number of points in the grid (see @eq-equationp2). Next, the bins that have a standard number of points less than a certain benchmark value are removed. After removing the hexagons that have insufficient data density, the hexagons with more substantial data representation are used to construct the 2D model.

@@

Furthermore, it is crucial to choose the benchmark value to remove low-density hexagons carefully. If we remove unnecessary bins, then it may result in long edges and an uneven 2D model. Therefore, instead of removing hexagons only identified by the benchmark value, we examine the standard number of points in the neighboring hexagons of the identified low-density hexagons. If the neighboring bins also have low counts, then only those bins will be removed. The other identified bins will be kept and used for constructing the 2D model along with the high-density bins.

The benchmark value for removing low-density hexagons has a range of $0$ and $1$. When analyzing how these benchmark values affect model performance, it's important to observe the change in Mean Squared Error (MSE) as the benchmark value increases. The MSE exhibits a gradual increase as the benchmark value progresses from $0$ to $1$. Assessing this rate of increase is crucial. If the increment is not substantial, the decision might lean towards retaining low-density hexagons.

@@

```{r}
#| echo: false
#| fig-cap: "Distribution of 2D Euclidean distances."
#| label: fig-stdctsScurve

cell_count_plot
```

@@

```{r}
#| echo: false
#| fig-cap: "(a) Full hexagonal grid with bin centroids, (b) Model constructed in 2D with long edges, and (c) Model constructed in 2D after removing the long edges."
#| label: fig-modelScurve
#| fig-pos: H

@@

```{r}
#| echo: false
#| fig-cap: "Distribution of 2D Euclidean distances."
#| label: fig-distScurve

distance_plot
```

@@

## UMAP
## Decide by looking at MSE plot
num_bins_x_s_curve <- 6
num_bins_y_s_curve <- 14
hex_size_s_curve <- 0.13

## Possible values for x_start
max_x_start <- min(umap_s_curve_scaled[["UMAP1"]]) + (sqrt(3) * hex_size_s_curve)
min_x_start <- min(umap_s_curve_scaled[["UMAP1"]]) - (sqrt(3) * hex_size_s_curve)

## Possible values for y_start
max_y_start <- min(umap_s_curve_scaled[["UMAP2"]]) + (1.5 * hex_size_s_curve)
min_y_start <- min(umap_s_curve_scaled[["UMAP2"]]) - (1.5 * hex_size_s_curve)

diff_start_df <- expand.grid(x_start = seq(min_x_start, max_x_start, by = 0.01), 
                             y_start = seq(min_y_start, max_y_start, by = 0.01))

vec <- stats::setNames(rep("", 4), c("x_start", "y_start", "num_non_empty_bins", "mse"))  ## Define column names

for (i in 1:NROW(diff_start_df)) {

  model_object <- fit_highd_model( training_data = training_data, 
                                   nldr_df_with_id = umap_s_curve_scaled, 
                                   x = "UMAP1", y = "UMAP2", 
                                   num_bins_x = num_bins_x_s_curve, 
                                   num_bins_y = num_bins_y_s_curve, 
                                   x_start = diff_start_df[i, 1], y_start = diff_start_df[i, 2], 
                                   buffer_x = NA, buffer_y = NA, 
                                   hex_size = hex_size_s_curve,
                                   is_rm_lwd_hex = FALSE, 
                                   benchmark_to_rm_lwd_hex = NA, 
                                   col_start_2d = "UMAP", 
                                   col_start_highd = "x")

  mse_df <- mse_df |>
    tibble::add_row(x_start = diff_start_df[i, 1], y_start = diff_start_df[i, 2],
                    num_non_empty_bins = NROW(centroid_df_training), 
                    mse = eval_list$mse)

## If same total number of bins occurred only select ones with minimum error
### Obtain duplicate bins
dupli_bins <- mse_df |> 
  dplyr::count(num_non_empty_bins) |> 
  dplyr::filter(n > 1) |> 
  dplyr::pull(num_non_empty_bins)

### Group split by duplicated bins
duplicate_df_list <- mse_df |>
  dplyr::filter(num_non_empty_bins %in% dupli_bins) |>
  dplyr::arrange(num_non_empty_bins) |>
  dplyr::group_split(num_non_empty_bins)

  dd <- duplicate_df_list[[i]] |>
    dplyr::filter(mse == min(duplicate_df_list[[i]]$mse))

### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df |>
  dplyr::filter(!(num_non_empty_bins %in% dupli_bins))

### Combine duplicated and not duplicated(corrected) bins dfs
mse_df <- dplyr::bind_rows(not_dupli_df, duplicate_df)

mse_plot <- ggplot(mse_df, aes(x = num_non_empty_bins,
                               y = mse
)) +
  geom_point() +
  geom_line() +
  # geom_vline(xintercept = 84, linetype="solid",
  #            color = "red", linewidth=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("number of non-empty bins")

mse_plot
```

@@

To assess how well our method captures and represents the underlying structure of the high-D data, Mean Squared Error (MSE) is used. When computing MSE, total model error (see @sec-summary) is divided by the number of observations to make it as a mean value (see @eq-equation9).

@@

@@

phate_gau_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tibble::as_tibble(PHATE_data_gau), x = "PHATE1", y = "PHATE2"))) |>
  dplyr::rename(c("PHATE1" = "scaled_PHATE1",
                  "PHATE2" = "scaled_PHATE2")) |>
  dplyr::mutate(ID = 1:NROW(PHATE_data_gau))

## Decide by looking at MSE plot
num_bins_x_tsne_gau <- 60
num_bins_y_tsne_gau <- 156
hex_size_tsne_gau <- 0.01
### non-empty 179

@@

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_tsne_gau$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
# ggplot(data = hex_grid, aes(x = x, y = y)) +
#   geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
#   geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red", size = 0.1) +
#   coord_fixed()

@@

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| warning: false
#| echo: false

## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:9] |>
  mutate(ID = 1:NROW(training_data_pbmc))

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, 
                                                                 x = "UMAP1", y = "UMAP2"))) |>
  dplyr::mutate(cell_label = umap_pbmc$cell_label) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.5) +
  coord_equal() +
  theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
  theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
        legend.key.height = unit(0.25, 'cm'),
        legend.key.width = unit(0.25, 'cm')) 
```

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?"
#| label: fig-umapauthor
#| fig-pos: H

umap_plot_pbmc 
```

<!-- MSE plot to find which is the effective number of bins to use-->
```{r}
#| warning: false
#| echo: false
#| message: false
